---
title: "Week 3 Answers"
author: "Jacob T. Fisher"
output: html_document
---

# Week 3 Answers
This file contains solutions for the Week 3 Exercise. Please don't consult this file until you've made a good-faith effort to solve the problems yourself.

## Loading the data and exploring the dataset

Load the data
```{r}
df <- read.csv("data/Cholesterol_R.csv")
```

Check out the first few rows of the dataset
```{r}
head(df)
```

Retrieve the column names
```{r}
names(df)
```

Retrieve the `Margarine` column
```{r}
df$Margarine
```

Retrieve the 10th element in the `Margarine` column
```{r}
df$Margarine[10]
```

Sum each row of the `Before`, `After4Weeks` and `After8Weeks` columns, storing the values in a column called `TotalCholesterol`. Tip. do not use `sum()` fuction, instead use a `+` sign
```{r}
df$TotalCholesterol <- df$Before + df$After4weeks + df$After8weeks
```

Add a new participant whose data are (19, 7.23, 6.86, 7.43, 0, 21.52).
```{r}
df[19,] <- c(19, 7.23, 6.86, 7.43, 0, 21.52)
```

Retrieve the new participant's data to see if the data are correctly recorded
```{r}
df[19,]
```

## Descriptive Analysis

Calculate the mean of the `TotalCholesterol` column
```{r}
mean(df$TotalCholesterol)
```

Calculate the variance of the `TotalCholesterol` column
Tip: Use the `var()` function
```{r}
var(df$TotalCholesterol)
```

Calculate the median of the `TotalCholesterol` column
```{r}
median(df$TotalCholesterol)
```

Generate a summary table of the `TotalCholesterol` column
```{r}
summary(df$TotalCholesterol)
```

Check the quantiles of the `TotalCholesterol` column
Tip: Use the `quantile()` function

```{r}
quantile(df$TotalCholesterol)
```

Use `sapply()` to compute the means of all variables except for `ID` and `Margarine`
```{r}
sapply(df[,-c(1, 6)], mean)
```

## Data Visualization

Create a histogram of the `TotalCholesterol` column with 10 breakpoints
```{r}
hist(df$TotalCholesterol, breaks = 10)
```

Create a boxplot of the `TotalCholesterol` column
```{r}
boxplot(df$TotalCholesterol)
```

Plot a bar graph of the `Margarine` column
```{r}
barplot(table(df$Margarine))
```

