---
title: "Data Processing and Management - COM 830 Tutorial"
author: "Jacob Fisher"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(haven)
library(knitr)
```

# Introduction

This tutorial covers essential data processing and management techniques
in R, with a focus on working with survey data from the World Values
Survey. We'll explore how to:

-   Load and examine SPSS data in R
-   Work with variable attributes and labels
-   Subset data efficiently using tidyverse principles
-   Recode variables
-   Calculate frequencies and percentages

# Loading and Examining SPSS Data

The World Values Survey data is stored in SPSS format (`.sav`). We'll
use the `haven` package, which is part of the tidyverse, to read this
data.

## Load SPSS data

```{r}
df <- read_spss("data/wvs_6 countries.sav")
```

## Understanding Variable Attributes

SPSS files contain additional metadata about variables, including:

-   Variable labels (descriptions)
-   Value labels (mapping of numeric codes to categorical meanings)
-   Variable types

Let's explore different ways to access this information.

### Method 1: View all attributes of a variable

```{r}
attributes(df$V2)
```

### Method 2: Access specific attributes

```{r}
attr(df$V2, 'label')  # Variable label
attr(df$V2, 'class')  # Variable class
attr(df$V2, 'labels') # Value labels
```

## Extracting Variable Information

For documentation and analysis purposes, we often need to extract all
variable labels and value labels:

### Get variable labels

Now we will extract information for all of the variables in the dataset.
Remember that `lapply()` is a function that:

-   Takes a list (or data frame) as input
-   Applies a function to each element of that list
-   Returns a list with the results

In simple terms, it's like saying "do this same thing to every piece of
my data." The `ldply()` function from the `plyr` package is used to
convert the list into a data frame.

```{r}
var_labels <- lapply(df, function(x) attributes(x)$label)
var_labels_df <- plyr::ldply(var_labels, data.frame)
names(var_labels_df) <- c("Var_Label", "Var_Name")
```

# Data Subsetting

There are multiple ways to subset data in R. We'll compare base R and
tidyverse approaches:

## Using Base R

Subset Australian males with all variables

```{r}
df_aus_male_all <- df[which(df$V2 == 36 & df$V240 == 1),]
```

Subset with specific columns

```{r}
df_aus_male <- df[which(df$V2 == 36 & df$V240 == 1), c("V2", "V240")]
```

## Using Tidyverse

The tidyverse approach is often more readable and maintainable:

Let's take a subset of the data that only contains US respondents (those
who have the country code `840`)

```{r}
df_usa <- df %>% 
  filter(V2 == 840)
```

Now we will select multiple columns from the dataset, then filter the
dataset to only contain US respondents. Remember that you can go back to
your `var_labels_df` to find the variable labels.

```{r}
df_subset <- df %>% 
  select(V2, V3, V4, V242) %>% 
  filter(V2 == 840)
```

# Variable Recoding

When working with survey data, we often need to recode variables into
new categories. Before recoding, it's important to examine the original
distribution:

## Check variable label and frequency distribution

Now we will check the variable label and frequency distribution of
`V242` (Age) in the dataset:

```{r}
attr(df$V242, "label")
freq <- table(df$V242)
prop.table(freq)
```

## Visualize the distribution

```{r}
hist(df$V242, main = "Distribution of Age", xlab = "Age")
```

## Creating groups based on the "Age" variable

Let's create age groups using a tidyverse approach. We'll define five
groups: 1-19, 20-29, 30-39, 40-49, and 50+. We will do this using the
`mutate()` function, which creates a new variable based on a condition.
The `case_when()` function allows us to specify multiple conditions.

```{r}
df_agegroup <- df %>%
  mutate(agegroup = case_when(
    V242 <= 19 ~ 1,
    V242 >= 20 & V242 < 30 ~ 2,
    V242 >= 30 & V242 < 40 ~ 3,
    V242 >= 40 & V242 < 50 ~ 4,
    V242 >= 50 ~ 5,
    TRUE ~ NA_real_
  ))
```

### Examining the new grouping

```{r}
table(df_agegroup$agegroup)
```

## Calculating Summary Statistics

The tidyverse provides a number of straightforward ways to calculate
frequencies, percentages, and cumulative percentages:

```{r}
age_summary <- df %>%
  count(V242) %>%
  mutate(
    percentage = n / sum(n) * 100,
    cumulative_percentage = cumsum(percentage)
  )
```

# Display the first few rows

```{r}
head(age_summary)
```

# Best Practices and Tips

Always examine your data before recoding: - Check variable labels and
value labels - Look at frequency distributions - Visualize when
appropriate

Document your code: - Include variable labels in comments - Explain the
purpose of transformations - Note any special cases or missing values

# Conclusion

This tutorial has covered the fundamental aspects of data processing and
management in R, with a focus on survey data. We've learned how to:

-   Work with SPSS data in R
-   Access and use variable metadata
-   Subset data efficiently
-   Recode variables
-   Calculate summary statistics
