---
title: "Week 5 Solutions: Data Processing and Missing Values"
author: "Your Name"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This document contains solutions to the Week 5 exercises on data processing and handling missing values.

# Part 1: Data Processing and Management

## Setup
### Load required packages
```{r}
library(tidyverse)
library(haven)
```

### Load the data
```{r}
df <- read_spss("data/wvs_6 countries.sav")
```

## Exercise 1: Working with Variable Labels

```{r}
# Check label and labels of V229
attr(df$V229, "label")  # Variable label
attr(df$V229, "labels")  # Value labels
```

## Exercise 2: Frequency Distributions

```{r}
# Generate frequency table for V229
table(df$V229, useNA = "ifany")
```

## Exercise 3: Subsetting Data - Unemployed

```{r}
# Create subset of unemployed respondents
df_unemployed <- df %>%
  filter(V229 == 7)  # Assuming 7 is the code for unemployed
```

## Exercise 4: Subsetting Data - Full Time Employment

```{r}
# Create subset of full-time employed
df_fulltime <- df %>%
  filter(V229 == 1)  # Assuming 1 is the code for full-time employment
```

## Exercise 5: Selecting Variables

```{r}
# Select specific variables from full-time employed dataset
df_subset <- df_fulltime %>%
  select(V1, V2, V3, V4, V229, V239)
```

## Exercise 6: Variable Information

```{r}
# Check label and value labels of V239
attr(df_subset$V239, "label")
attr(df_subset$V239, "labels")
```

## Exercise 7: Distribution Analysis

```{r}
# Check distribution of V239
table(df_subset$V239, useNA = "ifany")
```

## Exercise 8: Recoding Variables

```{r}
# Recode V239
df_subset <- df_subset %>%
  mutate(V239_recoded = case_when(
    V239 >= 1 & V239 <= 3 ~ 1,
    V239 >= 4 & V239 <= 6 ~ 2,
    TRUE ~ 3
  ))
```

## Exercise 9: Checking Recoding

```{r}
# Check frequency of recoded variable
table(df_subset$V239_recoded, useNA = "ifany")
```

# Part 2: Handling Missing Values

## Exercise 10: Variable Selection

```{r}
# Create subset with V9, V10, V11
df_missing <- df %>%
  select(V9, V10, V11)
```

## Exercise 11: Missing Value Check

```{r}
# Check for missing values
sapply(df_missing, function(x) sum(is.na(x)))
```

## Exercise 12: Impact of Missing Values

```{r}
# Count rows before and after removing missing values
original_rows <- nrow(df_missing)
complete_rows <- nrow(na.omit(df_missing))
cases_lost <- original_rows - complete_rows

print(paste("Original cases:", original_rows))
print(paste("Cases after removal:", complete_rows))
print(paste("Cases that would be lost:", cases_lost))
```

## Exercise 13: Mean Imputation

```{r}
# Replace missing values in V9 with mean
library(Hmisc)
df_missing$V9_mean_imputed <- impute(df_missing$V9, mean)
```

## Exercise 14: Comparing Means

```{r}
# Compare means
original_mean <- mean(df_missing$V9, na.rm = TRUE)
imputed_mean <- mean(df_missing$V9_mean_imputed)

print(paste("Original mean:", round(original_mean, 2)))
print(paste("Imputed mean:", round(imputed_mean, 2)))
```

## Exercise 15: Median Imputation

```{r}
# Replace missing values in V10 with median
df_missing$V10_median_imputed <- impute(df_missing$V10, median)
```

## Exercise 16: Comparing Results

```{r}
# Compare means
original_mean_v10 <- mean(df_missing$V10, na.rm = TRUE)
imputed_mean_v10 <- mean(df_missing$V10_median_imputed)

print(paste("Original mean:", round(original_mean_v10, 2)))
print(paste("Imputed mean:", round(imputed_mean_v10, 2)))
```

# Notes on Solutions

Key points to remember:
1. Always check variable labels and value labels before working with survey data
2. Use tidyverse functions for clear and readable code
3. Document the impact of missing value treatments
4. Validate results after recoding or imputation

Common pitfalls to avoid:
1. Forgetting to check for missing values
2. Not documenting value labels
3. Failing to verify recoding results
4. Not considering the impact of imputation on your analysis
