---
title: "Week 5 Exercises: Data Processing and Missing Values"
author: "Your Name"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this exercise, you'll practice working with the World Values Survey data, focusing on two main areas:
1. Data processing and management
2. Handling missing values

# Part 1: Data Processing and Management

First, let's set up our environment and load the necessary packages and data.

## Setup

### Load required packages
```{r}
# Your code here:
```

### Load the data

```{r}
# Your code here:
```

## Exercise 1: Working with Variable Labels
Check the label and labels (value labels) of variable V229.

```{r}
# Your code here:

```

## Exercise 2: Frequency Distributions
Generate a frequency table for V229.

```{r}
# Your code here:

```

## Exercise 3: Subsetting Data - Unemployed
Create a subset of the data containing only unemployed respondents.

```{r}
# Your code here:
```

## Exercise 4: Subsetting Data - Full Time Employment
Create a subset containing only people with full-time jobs.

```{r}
# Your code here:
```

## Exercise 5: Selecting Variables
Using the data from Exercise 4, create a subset containing only variables V1, V2, V3, V4, V229, and V239.

```{r}
# Your code here:
```

## Exercise 6: Variable Information
Using the data from Exercise 5, check the label and value labels of V239.

```{r}
# Your code here:
```

## Exercise 7: Distribution Analysis
Check the distribution of V239.

```{r}
# Your code here:
```

## Exercise 8: Recoding Variables
Recode V239 according to these rules:
- Values 1 to 3 become 1
- Values 4 to 6 become 2
- All other values become 3

Hint: Use `case_when()`

```{r}
# Your code here:
```

## Exercise 9: Checking Recoding
Check the frequency distribution of your recoded variable.

```{r}
# Your code here:
```

# Part 2: Handling Missing Values

## Exercise 10: Variable Selection
Create a subset of the data containing only variables V9, V10, and V11.

```{r}
# Your code here:
```

## Exercise 11: Missing Value Check
Check whether V9, V10, and V11 contain any missing values.

```{r}
# Your code here:
```

## Exercise 12: Impact of Missing Values
Using your subset from Exercise 10, determine how many cases would be deleted if you removed all rows with any missing values.

```{r}
# Your code here:
```

## Exercise 13: Mean Imputation
Replace missing values in V9 with its mean.

```{r}
# Your code here:
```

## Exercise 14: Comparing Means
Compare the mean of the original V9 with the mean of your new variable without missing values.

```{r}
# Your code here:
```

## Exercise 15: Median Imputation
Replace missing values in V10 with its median.

```{r}
# Your code here:
```

## Exercise 16: Comparing Results
Compare the mean of the original V10 with the mean of your new variable without missing values.

```{r}
# Your code here:
```
